---
title: "The US Opioid Crisis"
author: "Carmel Solomon"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---

```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```


## Introduction

The United States opioid crisis is one of the most devastating tragedies in modern history. It developed in three waves. The first wave began in the late 1990s when pharmaceutical companies marketed prescription opioid painkillers which lead to widespread overprescribing. Secondly, the second wave started in 2010 when people who were addicted to prescription opioids switched to cheaper heroin. The third and deadliest wave began around 2013 with the rise of illegally manufactured fentanyl. This increased the drug supply and caused deaths to surge dramatically.

This report uses time series analysis methods to examine the number of opioid overdose deaths in the US each year from 1999 until 2023. The dataset is sourced from the National Institute on Drug Abuse (NIDA), which gathers statistics based on data from the CDC WONDER database.

My aim for this analysis is to examine trends over time in  deaths related to opioid. I will apply Metas Prophet forecasting model to estimate how these deaths may change in the coming years if historical patterns continue.



## 1 Load and Prepare the Data

The annual opioid overdose death counts are entered directly using published figures from NIDA, which are compiled from the CDC WONDER database. Entering the data directly avoids issues that can occur when knitting external data files.

```{r, echo=FALSE}
# Annual opioid-involved overdose death counts, United States, 1999-2023
# Source: National Institute on Drug Abuse / CDC WONDER

opioidYears <- 1999:2023

opioidDeaths <- c(
  8048,  # 1999
  8407,  # 2000
  9496,  # 2001
  11920, # 2002
  12936, # 2003
  13756, # 2004
  14918, # 2005
  16651, # 2006
  18516, # 2007
  14800, # 2008
  15597, # 2009
  16651, # 2010
  17483, # 2011
  18893, # 2012
  23166, # 2013
  28647, # 2014
  33091, # 2015
  42249, # 2016
  47600, # 2017
  46802, # 2018
  49860, # 2019
  68630, # 2020
  80411, # 2021
  81806, # 2022
  79358  # 2023
)

opioidData <- data.frame(Year = opioidYears, Deaths = opioidDeaths)

# Display the first few rows
head(opioidData)
```

The data covers 25 years and shows how the opioid issue has increased over time.

---

## 2 Initial Visualisation


```{r}
plot(opioidData$Year, opioidData$Deaths,
     type = "l",
     col = "darkred",
     lwd = 2,
     main = "US Opioid-Involved Overdose Deaths (1999-2023)",
     xlab = "Year",
     ylab = "Number of Deaths")

points(opioidData$Year, opioidData$Deaths, pch = 16, col = "darkred")
```

The plot reveals a clear upward trend. Deaths remained relatively low through the early 2000s before accelerating from around 2013, coinciding with the rise of illicitly manufactured fentanyl. The slight decrease in 2023 suggests the trend is slowing down, although the numbers are still high.

---

## 3 Exploring Different Time Frames

The opioid crisis occured in three waves. Splitting the data by era reveals how the rate of growth changed over time.
```{r}
wave1Data <- subset(opioidData, Year <= 2009)
wave2Data <- subset(opioidData, Year >= 2010 & Year <= 2012)
wave3Data <- subset(opioidData, Year >= 2013)

plot(opioidData$Year, opioidData$Deaths,
     type = "n",
     main = "Opioid Deaths by Era",
     xlab = "Year",
     ylab = "Number of Deaths")

lines(wave1Data$Year, wave1Data$Deaths, col = "steelblue", lwd = 2)
lines(wave2Data$Year, wave2Data$Deaths, col = "orange", lwd = 2)
lines(wave3Data$Year, wave3Data$Deaths, col = "darkred", lwd = 2)

legend("topleft",
       legend = c("Wave 1: Prescription (1999-2009)",
                  "Wave 2: Heroin (2010-2012)",
                  "Wave 3: Fentanyl (2013-2023)"),
       col = c("steelblue", "orange", "darkred"), lwd = 2)
```

The fentanyl era produced by far the steepest increase. Deaths more than doubled between 2019 and 2021 alone.
```{r}
cat("Wave 1 average deaths per year:", round(mean(wave1Data$Deaths), 0), "\n")
cat("Wave 2 average deaths per year:", round(mean(wave2Data$Deaths), 0), "\n")
cat("Wave 3 average deaths per year:", round(mean(wave3Data$Deaths), 0), "\n")
```

The average annual death toll in the fentanyl era is more than four times higher than in the prescription era.

---

## 4 Exploratory Analysis

### 4.1 Basic Properties

```{r}
# Convert to a time series object for further analysis
opioidTimeSeries <- ts(opioidData$Deaths, start = 1999, frequency = 1)

# Check key properties
summary(opioidTimeSeries)
```

The data is annual, so there is no seasonal pattern. Deaths increased from 8,048 in 1999 to 81,806 in 2022, showing a large rise over time.


### 4.2 Understanding the Growth Trend

A simple linear regression helps estimate the rate of increase in overdose deaths over time.

```{r}
linearModel <- lm(Deaths ~ Year, data = opioidData)
summary(linearModel)

# Overlay the regression line on the plot
plot(opioidData$Year, opioidData$Deaths,
     type = "l",
     col = "darkred",
     lwd = 2,
     main = "Opioid Deaths with Linear Trend",
     xlab = "Year",
     ylab = "Number of Deaths")

abline(linearModel, col = "steelblue", lwd = 2, lty = 2)
legend("topleft", legend = c("Actual Deaths", "Linear Trend"),
       col = c("darkred", "steelblue"), lty = c(1, 2), lwd = 2)
```

On average, opioid overdose deaths increased by around 2,962 deaths per year over the study period. However, the actual data clearly accelerates beyond the straight line in later years, suggesting a model like Prophet is more suitable.


### 4.3 Autocorrelation Analysis
```{r}
acf(opioidTimeSeries, main = "ACF: Opioid Overdose Deaths")
pacf(opioidTimeSeries, main = "PACF: Opioid Overdose Deaths")
```

The ACF shows strong positive correlation, reflecting the upward trend. The PACF has a clear spike at lag 1, suggesting the previous year has the most influence.


---
## 5 Prophet Forecasting

### 5.1 Preparing Data for Prophet

```{r}
library(prophet)

# Create Prophet-compatible dataframe
prophetData <- data.frame(
  ds = as.Date(paste0(opioidYears, "-01-01")),
  y  = opioidDeaths
)

head(prophetData)
```

Metas Prophet requires data in a specific format with a date column named `ds` and a value column named `y`. Since our data is annual, dates are set to the 1st of January of each year.



### 5.2 Fitting the Prophet Model

```{r}
prophetModel <- prophet(
  prophetData,
  yearly.seasonality = FALSE,
  weekly.seasonality = FALSE,
  daily.seasonality = FALSE
)
```

The Prophet model is fitted to the data to cover the overall trend in opioid overdose deaths. Seasonality is not included because the data is annual and does not contain seasonal patterns.


### 5.3 Forecasting Future Values

```{r}
futureDates <- make_future_dataframe(prophetModel, periods = 5, freq = "year")
opioidForecast <- predict(prophetModel, futureDates)

plot(prophetModel, opioidForecast)
```

The model is used to generate forecasts for the next five years. These forecasts are based on the patterns observed in the historical data.


### 5.4 Forecast Table 
```{r}
forecastTable <- opioidForecast[
  (nrow(opioidForecast)-5+1):nrow(opioidForecast),
  c("ds", "yhat", "yhat_lower", "yhat_upper")
]

forecastTable$ds <- format(forecastTable$ds, "%Y")
colnames(forecastTable) <- c("Year", "Predicted Deaths", "Lower Bound", "Upper Bound")

forecastTable
```

The forecast table shows the predicted number of opioid overdose deaths for 2024–2028. The predicted values remain high and continue to increase, while the lower and upper bounds show uncertainty around these forecasts.


## Conclusion

This analysis used Meta’s Prophet model to examine trends in opioid overdose deaths in the United States between 1999 and 2023. The time series shows a strong upward trend, with a noticeable acceleration in recent years linked to the rise of synthetic opioids such as fentanyl.

The Prophet model successfully captures this trend and produces forecasts suggesting deaths could remain high in the coming years if current patterns continue. However, forecasting models rely on historical data and does not account for future policy changes. Despite this, the results highlight the persistence and severity of the opioid crisis.


*Data source: National Institute on Drug Abuse (NIDA), sourced from CDC WONDER. Available at: https://nida.nih.gov/research-topics/trends-statistics/overdose-death-rates*



